On exponential sums arising from the classical doubling method

### 2025-02-28 paper.tex
Calvin comments
Calvin changes

Hence for a torus character pair $(T',\theta')$ for $G$, we may conjugate $\mathbb{T}'$ to a fixed torus $\mathbb{T}$ and associate to $\theta$ to an element $x_{T,\theta}$ of $X^*(\mathbb{T})_{(\mathbb{Q}/\mathbb{Z})_{p'}}$ via the equality
\[
    exp(\langle x_{T,\theta},y\rangle) = \theta(N_{T(\mathbb{F}^n):T(\mathbb{F})}(y(\zeta_n)))
\]
for large enough $n$. This is well defined up to the Weyl group, and in fact there is a correspondence between $(T',\theta')$ pairs and $F$-stable Weyl group orbits in $X^*(\mathbb{T})_{(\mathbb{Q}/\mathbb{Z})_{p'}}$ [Coloarry 13.9].  By composing with \ref{eq:DL}, we associate to a torus character pair a Frobenius stable semisimple conjugacy class in $G^*$.

@@

For a torus theta pair $(T,\theta)$ with associated Frobenius conjugacy classes $w \in W(\mathbb{T})$ and $(s) \subset G^*$ and an embedding $\rho:G^* \to  GL_n$.  Fix a Frobenius stable maximal torus $\mathbb{T}^* \subset G^*$ and a Frobenius stable maximal torus $\mathbb{T} \subset GL_n$ which contains $\rho(\mathbb{T}^*)$. Let $\tilde{T}$ be the torus of $GL_n$ corresponding to the Frobenius conjugacy class of $\rho(w) \in W(\mathbb{T})$ and $\tilde{\theta}$ such that $(\rho(s)) = (s_{\tilde{T},\tilde{\theta}})$. Define
\[
    g_{T,\rho}(\theta,\psi,\chi) := \varepsilon_{\algebraicGroup{
    \tilde{T}}}\sum_{t\in \tilde{T}} \tilde{\theta}(t)\chi(det(t))\psi(tr(t)).
\]
The following lemma shows that this is well defined.
\begin{lemma}
\label{lemma:gl_invariance}
\begin{enumerate}[(i)]
	\item In the case $G = GL_n$ and $\rho$ is the standard representation, $g_{T,\rho}(\theta,\psi,\chi)$ depends only on the semisimple conjugacy class of $(s)$.
	\item $g_{T,\rho}(\theta,\psi,\chi)$ is well defined i.e. independent of choice of $\mathbb{T}^* \subset G^*$ and $\mathbb{T} \subset GL_n$.
\end{enumerate}
\end{lemma}
\begin{proof}
(i) In this case when $T = \Pi_{i} \mathbb{F}^{n_i}$ and $\theta = \Pi_{i} \alpha_i$, we recover the Gauss sum
\[
	g_{T,\rho}(\theta,\psi,\chi)  = \varepsilon_{\algebraicGroup{T}} \Pi_i \tau(\alpha_i \times (\chi\circ Nm),\psi) .
\]
By by the Hasse-Davenport relation depends only on the geometric conjugacy class of $(T,\theta)$ hence the semisimple conjugacy class of $(s_{T,\theta})$ \calvin{maybe write more or less?}. 
\\
(ii) For a Frobenius stable semisimple conjugacy class $(s)$ in $GL_n$ in $GL_n$, we define
\[
	g_{(s)}(\psi,\chi) := g_{T,\rho}(\theta,\psi,\chi) \textnormal{ for some } (T,\theta) \textnormal{ such that } (s_{T,\theta}) = (s)
\]
By (i) this is well defined. Then
\[
	g_{T,\rho}(\theta,\psi,\chi)  = g_{(\rho(s_{T,\theta}))}(\psi,\chi)
\]
which is independent of of choice of $\mathbb{T}^* \subset G^*$ and $\mathbb{T} \subset GL_n$.
\end{proof}

@@

The idea of the proof is that the formula from Theorem \ref{thm:thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters} relates $\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi}/R_{T,\theta}(1)$ to $g_{T,\rho}(\theta,\psi,\chi)$ for an appropriate $\rho$. We will need the following computation:
\begin{lemma}
\label{lem:cochar_computation}
    Define $\phi:(\mathbb{Q}/\mathbb{Z})_{p'}^n \to (\mathbb{Q}/\mathbb{Z})_{p'}^{2n} = X_*(GL_{2n})_{(\mathbb{Q}/\mathbb{Z})_{p'}}$ by $\phi(x) = (x,-x)$.
    \begin{enumerate}
        \item $x_{\mathbb{F}_n^\times \times \mathbb{F}_n^\times,\alpha \times \alpha^{-1}} = \phi(x_{\mathbb{F}_n,\alpha})$
        \item $x_{\mathbb{F}_{2n}^\times,\theta'}=\phi(x_{(\mathbb{F}_{2n})^{Nm=1},\theta})$
    \end{enumerate}
    where $\theta'(t) = \theta(t^{1-q^m})$.
\end{lemma}
\begin{proof}
Recall that $x_{T,\theta}$ is defined by for sufficiently large $m$ with $\zeta_m = 1/(q^m-1)$ under the fixed identification $\overline{\mathbb{F}}^\times \cong (\mathbb{Q}/\mathbb{Z})_{p'}$
\[
    exp(\langle x_{T,\theta},y\rangle) = \theta(y(Nm_{T(\mathbb{F}_m):T(\mathbb{F})}(y(\zeta_m))).
\]
Thus the result follows from the computations:\\
(i)
\begin{align*}
    exp(\langle\phi(x_{\mathbb{F}_n,\alpha}),(y_1,y_2)\rangle) &= exp(\langle x_{\mathbb{F}_n,\alpha},\phi^*(y_1,y_2)\rangle) \\
    &= \alpha\left(Nm_{mn:n}(y_1(\zeta_{mn})\right) \alpha\left(Nm_{mn:n}(y_2(\zeta_{mn})\right)^{-1} \\
    &= \left(\alpha\times\alpha^{-1}\right)\left(Nm_{mn:n}((y_1,y_2)(\zeta_{mn}))\right)
\end{align*}
(ii)
\begin{align*}
    exp(\langle\phi(x_{(\mathbb{F}_{2n})^{Nm=1},\theta}),(y_1,y_2)\rangle) &= exp(\langle x_{(\mathbb{F}_{2n})^{Nm=1},\theta},\phi^*(y_1,y_2)\rangle) \\
    &= \theta\left(\frac{Nm_{2mn:2n}(y_1(\zeta_{2mn})}{Nm_{2mn:2n}(y_1(\zeta_{2mn})^{q^n}}\right) \theta\left(\frac{Nm_{2mn:2n}(y_2(\zeta_{2mn})}{Nm_{2mn:2n}(y_2(\zeta_{2mn})^{q^n}}\right)^{-1} \\
    &= \theta'\left(Nm_{2mn:2n}((y_1,y_2)(\zeta_{mn}))\right)
\end{align*}
\end{proof}

>>>>

Hence for a torus character pair $(T,\theta)$ for $G$, we may conjugate $\mathbb{T}$ to a fixed torus $\mathbb{T}'$ and associate to $\theta$ a twisted Frobenius (twisted by an element of $W$ in the twisted Weyl group conjugacy class corresponding to $T$) stable element $x_{T,\theta}$ of $X^*(\mathbb{T}')_{(\mathbb{Q}/\mathbb{Z})_{p'}}$ via the equality
\[
    exp(\langle x_{T,\theta},y\rangle) = \theta(N_{T(\mathbb{F}^n):T(\mathbb{F})}(y(\zeta_n)))
\]
for large enough $n$. This is well defined up to the Weyl group, and in fact there is a correspondence between $(T,\theta)$ pairs and $F$-stable Weyl group orbits in $X^*(\mathbb{T}')_{(\mathbb{Q}/\mathbb{Z})_{p'}}$ [Corollary 13.9].  By composing with \ref{eq:DL}, we associate to a torus character pair a Frobenius stable semisimple conjugacy class in $G^*$.

@@

For a torus theta pair $(T,\theta)$ with associated Frobenius conjugacy classes $w \in W(\mathbb{T})$ and $(s) \subset G^*$ and an embedding $\rho:G^* \to  GL_n$.  Fix a Frobenius stable maximal torus $\mathbb{T}^* \subset G^*$ and a Frobenius stable maximal torus $\mathbb{T} \subset GL_n$ which contains $\rho(\mathbb{T}^*)$. Let $\tilde{T}$ be the torus of $GL_n$ corresponding to the Frobenius conjugacy class of $\rho(w) \in W(\mathbb{T})$ and $\tilde{\theta}$ such that $(\rho(s)) = (s_{\tilde{T},\tilde{\theta}})$. Define
\[
    g_{T,\rho}(\theta,\psi,\chi) := \varepsilon_{\algebraicGroup{
    \tilde{T}}}\sum_{t\in \tilde{T}} \tilde{\theta}(t)\chi(det(t))\psi(tr(t)).
\]
\calivn{check $\epsilon$ factors and Hasse-Davenport} The following lemma shows that this is well defined.
\begin{lemma}
\label{lemma:gl_invariance}
\begin{enumerate}[(i)]
	\item In the case $G = GL_n$ and $\rho$ is the standard representation, $g_{T,\rho}(\theta,\psi,\chi)$ depends only on the semisimple conjugacy class of $(s)$.
	\item $g_{T,\rho}(\theta,\psi,\chi)$ is well defined i.e. independent of choice of $\mathbb{T}^* \subset G^*$ and $\mathbb{T} \subset GL_n$.
\end{enumerate}
\end{lemma}
\begin{proof}
(i) In this case when $T = \Pi_{i} \mathbb{F}^{n_i}$ and $\theta = \Pi_{i} \alpha_i$, we recover the Gauss sum
\[
	g_{T,\rho}(\theta,\psi,\chi)  = \varepsilon_{\algebraicGroup{T}} \Pi_i \tau(\alpha_i \times (\chi\circ Nm),\psi) .
\]
By by the Hasse-Davenport relation depends only on the geometric conjugacy class of $(T,\theta)$ hence the semisimple conjugacy class of $(s_{T,\theta})$ \calvin{maybe write more or less?}. 
\\
(ii) For a Frobenius stable semisimple conjugacy class $(s)$ in $GL_n$ in $GL_n$, we define
\[
	g_{(s)}(\psi,\chi) := g_{T,\rho}(\theta,\psi,\chi) \textnormal{ for some } (T,\theta) \textnormal{ such that } (s_{T,\theta}) = (s)
\]
By (i) this is well defined. Then
\[
	g_{T,\rho}(\theta,\psi,\chi)  = g_{(\rho(s_{T,\theta}))}(\psi,\chi)
\]
which is independent of of choice of $\mathbb{T}^* \subset G^*$ and $\mathbb{T} \subset GL_n$.
\end{proof}

@@

The idea of the proof is that the formula from Theorem \ref{thm:thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters} relates $\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi}/R_{T,\theta}(1)$ to $g_{T,\rho}(\theta,\psi,\chi)$ for an appropriate $\rho$. We will need the following computation:
\begin{lemma}
\label{lem:cochar_computation}
    Define $\phi:(\mathbb{Q}/\mathbb{Z})_{p'}^n \to (\mathbb{Q}/\mathbb{Z})_{p'}^{2n} = X_*(GL_{2n})_{(\mathbb{Q}/\mathbb{Z})_{p'}}$ by $\phi(x) = (x,-x)$.
    \begin{enumerate}
        \item $x_{\mathbb{F}_n^\times \times \mathbb{F}_n^\times,\alpha \times \alpha^{-1}} = \phi(x_{\mathbb{F}_n,\alpha})$
        \item $x_{\mathbb{F}_{2n}^\times,\theta'}=\phi(x_{(\mathbb{F}_{2n})^{Nm=1},\theta})$
    \end{enumerate}
    where $\theta'(t) = \theta(t^{1-q^n})$.
\end{lemma}
\begin{proof}
Recall that $x_{T,\theta}$ is defined by for sufficiently large $m$ with $\zeta_m = 1/(q^m-1)$ under the fixed identification $\overline{\mathbb{F}}^\times \cong (\mathbb{Q}/\mathbb{Z})_{p'}$
\[
    exp(\langle x_{T,\theta},y\rangle) = \theta(y(Nm_{T(\mathbb{F}_m):T(\mathbb{F})}(y(\zeta_m))).
\]
Thus the result follows from the computations:\\
(i)
\begin{align*}
    exp(\langle\phi(x_{\mathbb{F}_n,\alpha}),(y_1,y_2)\rangle) &= exp(\langle x_{\mathbb{F}_n,\alpha},\phi^*(y_1,y_2)\rangle) \\
    &= \alpha\left(Nm_{mn:n}(y_1(\zeta_{mn})\right) \alpha\left(Nm_{mn:n}(y_2(\zeta_{mn})\right)^{-1} \\
    &= \left(\alpha\times\alpha^{-1}\right)\left(Nm_{mn:n}((y_1,y_2)(\zeta_{mn}))\right)
\end{align*}
(ii)
\begin{align*}
    exp(\langle\phi(x_{(\mathbb{F}_{2n})^{Nm=1},\theta}),(y_1,y_2)\rangle) &= exp(\langle x_{(\mathbb{F}_{2n})^{Nm=1},\theta},\phi^*(y_1,y_2)\rangle) \\
    &= \theta\left(\frac{Nm_{2mn:2n}(y_1(\zeta_{2mn})}{Nm_{2mn:2n}(y_1(\zeta_{2mn})^{q^n}}\right) \theta\left(\frac{Nm_{2mn:2n}(y_2(\zeta_{2mn})}{Nm_{2mn:2n}(y_2(\zeta_{2mn})^{q^n}}\right)^{-1} \\
    &= \theta'\left(Nm_{2mn:2n}((y_1,y_2)(\zeta_{mn}))\right)
\end{align*}
\end{proof}
